---
title: "Standard Operating Procedure: Processing Commercial Wearable Step-Count Data"
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

---

# Overview

This document implements a transparent, reproducible standard operating procedure
(SOP) for transforming epoch-level step-count data from commercial Garmin devices
into participant-level analytic variables. The pipeline standardizes timestamps,
reconstructs daily epoch grids, infers wear time from observed step patterns, and
applies a prespecified valid-day threshold to generate daily and participant-level
summaries.

The pipeline is designed for settings where device-generated wear-time indicators
are unavailable. All operational definitions ‚ÄĒ epoch length, wear-time inference
rules, valid-day threshold, and quality-control criteria ‚ÄĒ are declared in a single
configuration block and propagated throughout, so that any parameter change
requires editing only one location.

---

# Configuration

```{r config}
# ‚ĒÄ‚ĒÄ File paths ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ
PATH_STEPS   <- "path/to/raw_step_count_data.xlsx"
PATH_FILLED  <- "path/to/output/filled_step_counts.csv"       # optional save
PATH_SUMMARY <- "path/to/output/participant_step_summary.csv" # optional save

# ‚ĒÄ‚ĒÄ Processing parameters ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ‚ĒÄ
EPOCH_MIN        <- 15   # epoch length in minutes
MAX_GAP_EPOCHS   <- 100  # consecutive missing epochs >= this are treated as non-wear
VALID_DAY_HR_PRI <- 10   # primary valid-day threshold (hours); consistent with NHANES convention
VALID_DAY_HR_SEN <- 6    # sensitivity valid-day threshold (hours)
MIN_VALID_DAYS   <- 4    # minimum valid days to retain a participant in summaries
```

---

# Step 1: Load and Inspect Raw Data

Raw data are imported from the device platform export. Timestamps are standardized
and the time zone offset is recorded. The raw file is expected to contain one row
per epoch with at minimum: a participant identifier, epoch start time, epoch end
time, and step count value.

```{r load}
step_data_raw <- read_excel(PATH_STEPS) %>%
  rename(
    participant_id   = `User Name`,
    start_time       = `Start Time`,
    end_time         = `End Time`,
    time_zone_offset = `Time Zone Offset`,
    value            = Value
  )

cat("Participants in raw data:", n_distinct(step_data_raw$participant_id), "\n")
cat("Total epoch records:     ", nrow(step_data_raw), "\n")
cat("Date range:              ",
    format(min(as.Date(step_data_raw$start_time))), "to",
    format(max(as.Date(step_data_raw$start_time))), "\n")
cat("Time zone offsets present:",
    paste(unique(step_data_raw$time_zone_offset), collapse = ", "), "\n")
```

---

# Step 2: Standardize Timestamps and Remove Irregular Epochs

Duplicate records sharing the same participant and start time are removed, retaining
the longer epoch. Epochs exceeding the nominal epoch length ‚ÄĒ which may reflect
device export artifacts or non-wear segments ‚ÄĒ are excluded.

```{r clean}
stepsall <- step_data_raw %>%
  mutate(
    length = as.numeric(difftime(end_time, start_time, units = "mins")),
    date   = as.Date(start_time)
  ) %>%
  arrange(desc(length)) %>%
  distinct(participant_id, start_time, .keep_all = TRUE) %>%
  filter(length <= EPOCH_MIN) %>%
  select(participant_id, start_time, end_time, date, length, value)

cat("Epoch records after cleaning:", nrow(stepsall), "\n")
cat("Participants retained:        ", n_distinct(stepsall$participant_id), "\n")

non_standard <- stepsall %>% filter(length != EPOCH_MIN)
if (nrow(non_standard) > 0) {
  message(nrow(non_standard), " epochs with non-standard length remaining:")
  print(non_standard)
} else {
  message("All retained epochs are exactly ", EPOCH_MIN, " minutes.")
}
```

---

# Step 3: Reconstruct Daily Epoch Grids and Infer Wear Time

## Operational definitions

**Epoch:** The smallest unit of time over which step counts are recorded. In this
implementation, step counts are aggregated in fixed EPOCH_MIN-minute intervals.
All processing is conducted at the epoch level prior to daily aggregation.

**Wear time:** Because direct wear-time indicators are unavailable, wear time is
operationally inferred from observed step patterns. Within each calendar day, the
wear window is defined as the interval between the first and last detected non-zero
step epoch. Zero-step epochs falling within this window are retained and treated as
potential inactivity. Gaps of MAX_GAP_EPOCHS or more consecutive missing epochs
within the wear window are treated as non-wear and excluded from imputation.

**Valid day:** A calendar day with at least VALID_DAY_HR_PRI hours of inferred
wear time under the primary threshold, or VALID_DAY_HR_SEN hours under the
sensitivity threshold.

## Helper functions

```{r helpers}

#' Identify runs of consecutive integers of length >= min_run
#'
#' @param x       Integer vector (need not be sorted)
#' @param min_run Minimum run length to flag
#' @return List: `sequences` (list of integer vectors), `count` (number of runs found)
find_consecutive_runs <- function(x, min_run) {
  x <- sort(x)
  if (length(x) == 0) return(list(sequences = list(), count = 0L))
  grp       <- cumsum(c(1L, diff(x) != 1L))
  runs      <- split(x, grp)
  long_runs <- runs[lengths(runs) >= min_run]
  list(sequences = long_runs, count = length(long_runs))
}


#' Insert a single zero-step epoch into a participant-day data frame
#'
#' @param df             Data frame for one participant-day (must contain `index` column)
#' @param missing_index  Epoch index to insert
#' @param subj           participant_id string
#' @param date_subj      Date object
#' @param time_intervals POSIXct vector of all epoch start times for that day
#' @return Data frame with the new zero-step row inserted, sorted by index
insert_zero_epoch <- function(df, missing_index, subj, date_subj, time_intervals) {
  new_row                <- df[1L, ]
  new_row[]              <- NA
  new_row$participant_id <- subj
  new_row$start_time     <- time_intervals[missing_index]
  new_row$end_time       <- time_intervals[missing_index + 1L]
  new_row$date           <- as.Date(date_subj)
  new_row$length         <- EPOCH_MIN
  new_row$value          <- 0L
  new_row$index          <- missing_index
  bind_rows(df, new_row) %>% arrange(index)
}


#' Fill zero-step epochs within a participant-day wear window
#'
#' Missing epochs between the first and last observed epoch are imputed as zero
#' steps, except for runs of >= MAX_GAP_EPOCHS consecutive missing epochs, which
#' are treated as non-wear and left unfilled.
#'
#' @param subj       participant_id string
#' @param date_subj  Date object
#' @return List: `data` (filled data frame), `n_nonwear_gaps` (count of long gaps skipped)
fill_day <- function(subj, date_subj) {
  day_df <- stepsall %>%
    filter(participant_id == subj, date == date_subj) %>%
    distinct()

  time_intervals <- seq(
    from = as.POSIXct(paste(date_subj, "00:00:00")),
    to   = as.POSIXct(paste(date_subj, "23:45:00")),
    by   = paste(EPOCH_MIN, "mins")
  )

  day_df$index <- match(day_df$start_time, time_intervals)
  valid_idx    <- day_df$index[!is.na(day_df$index)]
  all_idx      <- seq(min(valid_idx), max(valid_idx))
  missing_idx  <- setdiff(all_idx, valid_idx)

  if (length(missing_idx) <= 1L) {
    n_nonwear_gaps <- 0L
  } else {
    runs           <- find_consecutive_runs(missing_idx, MAX_GAP_EPOCHS)
    n_nonwear_gaps <- runs$count
    missing_idx    <- setdiff(missing_idx, unlist(runs$sequences))
  }

  for (idx in missing_idx) {
    day_df <- insert_zero_epoch(day_df, idx, subj, date_subj, time_intervals)
  }

  list(data = day_df, n_nonwear_gaps = n_nonwear_gaps)
}
```

## Apply across all participant-days

Days with only a single observed epoch are excluded before filling, as a wear
window cannot be defined from a single point.

```{r impute}
subj_dates <- stepsall %>%
  group_by(participant_id, date) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1L) %>%
  mutate(date = as.Date(date))

results_list <- vector("list", nrow(subj_dates))
for (i in seq_len(nrow(subj_dates))) {
  results_list[[i]] <- fill_day(subj_dates$participant_id[i], subj_dates$date[i])
}

filled_data_all <- bind_rows(lapply(results_list, `[[`, "data"))

cat("Participant-days processed:", nrow(subj_dates), "\n")
cat("Participants:              ", n_distinct(filled_data_all$participant_id), "\n")

# Optional: save filled epoch data
# write.csv(filled_data_all, PATH_FILLED, row.names = FALSE)
```

---

# Step 4: Apply Valid-Day Threshold and Generate Summaries

Participant-level metrics ‚ÄĒ mean daily steps, within-person standard deviation,
coefficient of variation, and number of valid days ‚ÄĒ are derived from the valid-day
set under both thresholds. Participants with fewer than MIN_VALID_DAYS valid
days are excluded from summaries.

```{r summarise}
epochs_per_hr <- 60L / EPOCH_MIN

#' Compute participant-level step-count summaries for a given valid-day threshold
#'
#' @param threshold_hr  Minimum inferred wear time (hours) to classify a day as valid
#' @return Participant-level data frame: mean_steps, sd_steps, cv_steps, valid_days
summarise_steps <- function(threshold_hr) {
  filled_data_all %>%
    group_by(participant_id, date) %>%
    summarise(
      wear_time_hr = n() / epochs_per_hr,
      total_steps  = sum(value, na.rm = TRUE),
      .groups      = "drop"
    ) %>%
    filter(wear_time_hr >= threshold_hr) %>%
    group_by(participant_id) %>%
    summarise(
      mean_steps = mean(total_steps, na.rm = TRUE),
      sd_steps   = sd(total_steps,   na.rm = TRUE),
      cv_steps   = sd_steps / mean_steps,
      valid_days = n(),
      .groups    = "drop"
    ) %>%
    filter(valid_days >= MIN_VALID_DAYS)
}

summary_pri <- summarise_steps(VALID_DAY_HR_PRI)
summary_sen <- summarise_steps(VALID_DAY_HR_SEN)

cat("Participants with ‚Č•", MIN_VALID_DAYS, "valid days:\n")
cat("  Primary threshold    (‚Č•", VALID_DAY_HR_PRI, "h):", nrow(summary_pri), "\n")
cat("  Sensitivity threshold (‚Č•", VALID_DAY_HR_SEN, "h):", nrow(summary_sen), "\n")

# Optional: save primary participant-level summary
# write.csv(summary_pri, PATH_SUMMARY, row.names = FALSE)
```

---

# Step 5: Quality Control

Descriptive summaries and visualizations are generated prior to downstream analysis
to identify implausible or irregular step-count patterns that could reflect device
malfunction, data corruption, or substantial non-wear.

```{r qc_setup}
# Daily-level data under the primary threshold, used throughout QC
daily_pri <- filled_data_all %>%
  group_by(participant_id, date) %>%
  summarise(
    wear_time_hr = n() / epochs_per_hr,
    total_steps  = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(wear_time_hr >= VALID_DAY_HR_PRI)

theme_sop <- function() {
  theme_minimal() +
    theme(plot.title = element_text(face = "bold"))
}
```

## Wear-time distribution on valid days

```{r qc_weartime, fig.width=7, fig.height=4}
ggplot(daily_pri, aes(x = wear_time_hr)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  labs(
    title = paste0("Daily Wear Time on Valid Days (‚Č•", VALID_DAY_HR_PRI, "h threshold)"),
    x     = "Inferred wear time (hours)",
    y     = "Participant-days"
  ) +
  theme_sop()
```

## Daily step-count distribution on valid days

```{r qc_daily_steps, fig.width=7, fig.height=4}
ggplot(daily_pri, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "steelblue", color = "white") +
  labs(
    title = paste0("Daily Step Counts on Valid Days (‚Č•", VALID_DAY_HR_PRI, "h threshold)"),
    x     = "Daily step count",
    y     = "Participant-days"
  ) +
  theme_sop()
```

## Participant-level mean daily steps

```{r qc_participant_steps, fig.width=7, fig.height=4}
sample_median <- median(summary_pri$mean_steps)

ggplot(summary_pri, aes(x = mean_steps)) +
  geom_histogram(binwidth = 1000, fill = "steelblue", color = "white") +
  geom_vline(xintercept = sample_median, linetype = "dashed", color = "firebrick") +
  labs(
    title    = "Participant-Level Mean Daily Steps",
    subtitle = paste0("Primary threshold (‚Č•", VALID_DAY_HR_PRI,
                      "h); dashed line = sample median (",
                      round(sample_median), " steps)"),
    x = "Mean daily steps",
    y = "Participants"
  ) +
  theme_sop()
```

## Per-participant daily step boxplots

Per-participant boxplots of daily step counts provide a visual quality check for
outlying days or participants with implausible patterns.

```{r qc_boxplots, fig.width=10, fig.height=5}
# Order participants by their mean daily steps for readability
participant_order <- summary_pri %>%
  arrange(mean_steps) %>%
  pull(participant_id)

daily_pri %>%
  mutate(participant_id = factor(participant_id, levels = participant_order)) %>%
  ggplot(aes(x = participant_id, y = total_steps)) +
  geom_boxplot(fill = "steelblue", color = "grey30",
               outlier.size = 1, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 18, size = 2, color = "firebrick") +
  labs(
    title    = "Daily Step Counts by Participant (sorted by mean)",
    subtitle = "Boxes = IQR and median; diamonds = mean",
    x        = "Participant (sorted by mean daily steps)",
    y        = "Daily steps"
  ) +
  theme_sop() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

---

# Step 6: Sensitivity Analysis ‚ÄĒ Valid-Day Threshold

Participant-level estimates are compared across the primary (VALID_DAY_HR_PRI h)
and sensitivity (VALID_DAY_HR_SEN h) valid-day thresholds to verify that the
choice of threshold does not materially influence derived step-count summaries.

```{r sensitivity_threshold}
# Participants retained under both thresholds
compare_thresholds <- summary_sen %>%
  rename(mean_steps_sen = mean_steps, sd_steps_sen = sd_steps,
         cv_steps_sen   = cv_steps,   valid_days_sen = valid_days) %>%
  inner_join(
    summary_pri %>%
      rename(mean_steps_pri = mean_steps, sd_steps_pri = sd_steps,
             cv_steps_pri   = cv_steps,   valid_days_pri = valid_days),
    by = "participant_id"
  ) %>%
  mutate(
    days_lost       = valid_days_sen - valid_days_pri,
    pct_days_lost   = if_else(valid_days_sen > 0, days_lost / valid_days_sen, NA_real_),
    diff_mean_steps = mean_steps_pri - mean_steps_sen,
    diff_cv         = cv_steps_pri   - cv_steps_sen
  )

dropped_by_primary <- anti_join(summary_sen, summary_pri, by = "participant_id")

cat("Participants under ‚Č•", VALID_DAY_HR_SEN, "h (sensitivity):", nrow(summary_sen), "\n")
cat("Participants under ‚Č•", VALID_DAY_HR_PRI, "h (primary):    ", nrow(summary_pri), "\n")
cat("Dropped by primary threshold:            ", nrow(dropped_by_primary), "\n\n")

cat("Days lost (strict ‚ąí lenient) ‚ÄĒ median [min, max]:\n  ",
    median(compare_thresholds$days_lost), " [",
    min(compare_thresholds$days_lost), ", ",
    max(compare_thresholds$days_lost), "]\n\n", sep = "")

cat("Proportion of valid days lost ‚ÄĒ median [min, max]:\n  ",
    round(median(compare_thresholds$pct_days_lost), 3), " [",
    round(min(compare_thresholds$pct_days_lost), 3), ", ",
    round(max(compare_thresholds$pct_days_lost), 3), "]\n\n", sep = "")

cat("őĒ mean daily steps (primary ‚ąí sensitivity) ‚ÄĒ median [min, max]:\n  ",
    round(median(compare_thresholds$diff_mean_steps), 1), " [",
    round(min(compare_thresholds$diff_mean_steps), 1), ", ",
    round(max(compare_thresholds$diff_mean_steps), 1), "]\n\n", sep = "")

tibble(
  Threshold           = c(paste0("‚Č•", VALID_DAY_HR_SEN, "h (sensitivity)"),
                          paste0("‚Č•", VALID_DAY_HR_PRI, "h (primary)")),
  n                   = c(nrow(summary_sen), nrow(summary_pri)),
  `Median valid days` = c(median(compare_thresholds$valid_days_sen),
                          median(compare_thresholds$valid_days_pri)),
  `Median mean steps` = c(round(median(compare_thresholds$mean_steps_sen)),
                          round(median(compare_thresholds$mean_steps_pri))),
  `Median CV`         = c(round(median(compare_thresholds$cv_steps_sen), 3),
                          round(median(compare_thresholds$cv_steps_pri), 3))
) %>% print()
```

```{r sensitivity_threshold_plot, fig.width=7, fig.height=4}
compare_thresholds %>%
  select(participant_id, mean_steps_sen, mean_steps_pri) %>%
  pivot_longer(
    cols      = c(mean_steps_sen, mean_steps_pri),
    names_to  = "threshold",
    values_to = "mean_steps"
  ) %>%
  mutate(
    threshold = factor(threshold,
                       levels = c("mean_steps_sen", "mean_steps_pri"),
                       labels = c(paste0("‚Č•", VALID_DAY_HR_SEN, "h (sensitivity)"),
                                  paste0("‚Č•", VALID_DAY_HR_PRI, "h (primary)")))
  ) %>%
  ggplot(aes(x = threshold, y = mean_steps, group = participant_id)) +
  geom_line(color = "grey75", linewidth = 0.4) +
  geom_point(aes(color = threshold), size = 2, alpha = 0.75) +
  scale_color_manual(values = c("steelblue", "firebrick"), guide = "none") +
  labs(
    title    = "Mean Daily Steps by Valid-Day Threshold",
    subtitle = paste0("Each line = one participant (n = ",
                      nrow(compare_thresholds), " common to both thresholds)"),
    x = "Valid-day threshold",
    y = "Mean daily steps"
  ) +
  theme_sop()
```

---

# Participant-Level Summary

Final participant-level step-count estimates under the primary threshold.

```{r final_summary}
cat("‚ĒÄ‚ĒÄ Participant-level step-count summary",
    "(primary threshold ‚Č•", VALID_DAY_HR_PRI, "h) ‚ĒÄ‚ĒÄ\n")
cat("  n:                                    ", nrow(summary_pri), "\n")
cat("  Valid days ‚ÄĒ median [min, max]:       ",
    median(summary_pri$valid_days), " [",
    min(summary_pri$valid_days), ", ",
    max(summary_pri$valid_days), "]\n", sep = "")
cat("  Mean daily steps ‚ÄĒ median (Q1, Q3):  ",
    round(median(summary_pri$mean_steps)), " (",
    round(quantile(summary_pri$mean_steps, 0.25)), ", ",
    round(quantile(summary_pri$mean_steps, 0.75)), ")\n", sep = "")
cat("  Within-person SD ‚ÄĒ median (Q1, Q3):  ",
    round(median(summary_pri$sd_steps)), " (",
    round(quantile(summary_pri$sd_steps, 0.25)), ", ",
    round(quantile(summary_pri$sd_steps, 0.75)), ")\n", sep = "")
cat("  Coefficient of variation ‚ÄĒ median:    ",
    round(median(summary_pri$cv_steps), 3), "\n", sep = "")
```
